---
title: "Regression Results"
author: "Suntiparp Somsak"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: false
    self_contained: true
    css: |
      body { font-family: "Georgia", serif; font-size: 14px; color: #222; max-width: 960px; margin: auto; padding: 2em; }
      h3 { font-size: 1.05em; font-weight: bold; margin-top: 2em; margin-bottom: 0.3em; color: #333; }
      p.note { font-size: 0.82em; color: #555; line-height: 1.6; margin-top: 0.6em; }
      .table { font-size: 0.88em; }
      .table thead th { background-color: #1D6FA4; color: white; border-bottom: none; }
      .table tbody td { vertical-align: middle; }
      .table-striped tbody tr:nth-of-type(odd) { background-color: #f7fafd; }
      hr { border: none; border-top: 1px solid #ddd; margin: 2.5em 0; }
---

```{r setup, include=FALSE}

# 09_visualizations.Rmd
# Regression result tables: zero-dose model, recovery model,
# weather model, and model details
# Created: July 2026 | Last Updated: 03/07/2026

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(tidyverse)
library(knitr)
library(kableExtra)

#--- Paths ---
home    <- "C:/Users/HP/Documents/GitHub/datharm-placement"
reg_dir <- file.path(home, "03_output/03_regression")

#--- Parse modelsummary txt ---
parse_ms_txt <- function(path) {
  if (!file.exists(path)) return(NULL)
  lines <- readLines(path, warn = FALSE)
  rows  <- lines[str_detect(lines, "\\|") &
                 !str_detect(lines, "^[+=]+$") &
                 !str_detect(lines, "^\\+[-=+]+\\+$")]
  map_dfr(rows, function(r) {
    cells <- str_split(r, "\\|")[[1]] %>% str_trim() %>% .[. != ""]
    if (length(cells) >= 2) tibble(term = cells[1], values = list(cells[-1]))
    else NULL
  })
}

ev <- function(parsed, label, col = 1) {
  if (is.null(parsed)) return(NA_character_)
  row <- parsed %>% filter(str_detect(term, fixed(label)))
  if (nrow(row) == 0 || length(row$values[[1]]) < col) return(NA_character_)
  str_trim(row$values[[1]][col])
}

fb <- function(live, hard) if (is.na(live) || live == "") hard else live

ma <- parse_ms_txt(file.path(reg_dir, "01_model_a_zerodose_predictors.txt"))
mb <- parse_ms_txt(file.path(reg_dir, "02_model_b_tracing_effectiveness.txt"))

#--- Helper: format coef + 95% CI as two-line cell ---
ci_cell <- function(coef, se_num, stars = "") {
  se   <- as.numeric(se_num)
  lo   <- round(as.numeric(coef) - 1.96 * se, 3)
  hi   <- round(as.numeric(coef) + 1.96 * se, 3)
  coef_fmt <- paste0(coef, stars)
  ci_fmt   <- paste0("<small style='color:#666'>[", lo, ", ", hi, "]</small>")
  paste0(coef_fmt, "<br>", ci_fmt)
}

#--- Zero-dose model ---
a1_dist  <- ci_cell("0.043",  "0.023", "*")
a2_dist  <- ci_cell("0.627",  "0.199", "***")
a1_age   <- ci_cell("0.003",  "0.001", "**")
a2_age   <- ci_cell("0.061",  "0.008", "***")
a1_sex   <- ci_cell("0.041",  "0.032", "")
a2_sex   <- ci_cell("0.040",  "0.039", "")
a1_n     <- fb(ev(ma,"Num.Obs",1), "69,366")
a2_n     <- fb(ev(ma,"Num.Obs",2), "69,366")
a1_r2    <- fb(ev(ma,"R2 ",1),     "0.061")
a2_r2    <- fb(ev(ma,"R2 ",2),     "0.220")

#--- Recovery model ---
b1_sms   <- ci_cell("0.444",  "0.145", "***")
b1_dist  <- ci_cell("-0.051", "0.061", "")
b1_age   <- ci_cell("-0.005", "0.003", "")
b1_lag   <- ci_cell("-0.001", "0.001", "")
b1_n     <- fb(ev(mb,"Num.Obs",1), "16,132")
b1_r2    <- fb(ev(mb,"R2 ",1),     "0.117")

```

### Table 1. &nbsp; Zero-dose model and recovery model

```{r tab-zd-rec}

df <- tribble(
  ~term,                                    ~c1,      ~c2,      ~c3,            ~c4,
  "Distance to health facility (km)",       a1_dist,  a2_dist,  b1_dist,        b1_dist,
  "Age at registration / tracing (months)", a1_age,   a2_age,   b1_age,         b1_age,
  "Female (ref: male)",                     a1_sex,   a2_sex,   "—",            "—",
  "SMS / phone contact (ref: home visit)",  "—",      "—",      b1_sms,         b1_sms,
  "Days since last facility visit",         "—",      "—",      b1_lag,         b1_lag,
  "<em>N</em>",                             a1_n,     a2_n,     "40,236",       b1_n,
  "<em>R²</em>",                            a1_r2,    a2_r2,    "0.117",        b1_r2
)

df %>%
  kbl(
    col.names = c("", "(1)", "(2)", "(3)", "(4)"),
    align     = c("l","r","r","r","r"),
    escape    = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped","hover","condensed","bordered"),
    full_width = TRUE, font_size = 13
  ) %>%
  add_header_above(
    c(" " = 1,
      "Zero-dose model" = 2,
      "Recovery model"  = 2),
    background = c("white", "#1D6FA4", "#BA7517"),
    color      = c("white", "white",   "white"),
    bold       = TRUE
  ) %>%
  add_header_above(
    c(" " = 1,
      "Primary definition" = 1,
      "Strict definition"  = 1,
      "Full sample"        = 1,
      "Lag-time subset"    = 1),
    italic = TRUE, line = FALSE, font_size = 11
  ) %>%
  column_spec(1, width = "22em") %>%
  column_spec(2:5, width = "10em") %>%
  row_spec(5, extra_css = "border-bottom: 1px solid #ccc;") %>%
  row_spec(6:7, extra_css = "font-style: italic; border-top: 1px solid #ccc;")

```

<p class="note">
**Notes.** Estimates on the log-odds scale; 95% confidence intervals in brackets. Standard errors clustered at the ward level (zero-dose model) and LGA level (recovery model). The zero-dose primary definition uses the MCHTrack penta-ZD flag (children 12–23 months, no recorded first pentavalent dose); the strict definition requires no vaccine contact of any kind in facility records and cannot detect off-network vaccination. Recovery is defined as confirmed return to MCHTrack services (*yes_ok* only); the lag-time subset is restricted to records with a prior facility visit date on file. Rimi LGA excluded throughout. &nbsp; &#42;&nbsp;p&nbsp;&lt;&nbsp;0.10 &nbsp; &#42;&#42;&nbsp;p&nbsp;&lt;&nbsp;0.05 &nbsp; &#42;&#42;&#42;&nbsp;p&nbsp;&lt;&nbsp;0.01
</p>

<hr>

### Table 2. &nbsp; Weather model

```{r tab-weather}

wx <- tribble(
  ~Spec, ~Variable,       ~Measure,                                                ~Coef_CI,                                                            ~Panel,    ~FE,                       ~N,      ~R2,
  "W1",  "Precipitation", "Days per month above 38°C (monthly)",                   ci_cell("-0.006","0.002","***"),                                      "Monthly", "LGA",                     "24",    "0.942",
  "W2",  "Precipitation", "Daily precipitation anomaly (% of long-term average)",  ci_cell("-0.006","0.002",""),                                         "Daily",   "LGA, DOW, month-year",    "1,216", "0.698",
  "W3",  "Heat",          "Extreme heat day (UTCI ≥ 38°C, binary)",                ci_cell("-0.056","0.089",""),                                         "Daily",   "LGA, DOW, month-year",    "1,216", "0.698",
  "W4",  "Heat",          "Daytime mean UTCI (°C, centred)",                       ci_cell("0.019","0.014",""),                                          "Daily",   "LGA, DOW, month-year",    "1,216", "0.698",
  "W5",  "Vegetation",    "NDVI seasonal level (vim, centred)",                    ci_cell("0.036","0.193",""),                                          "Monthly", "LGA",                     "40",    "0.910",
  "W6",  "Vegetation",    "NDVI within-baseline anomaly (viq percentile, centred)",ci_cell("-0.053","0.016","***"),                                      "Monthly", "LGA, month-year",         "40",    "0.989"
)

wx %>%
  mutate(Variable = if_else(duplicated(Variable), "", Variable)) %>%
  kbl(
    col.names = c("", "Variable", "Exposure measure",
                  "Coef. [95% CI]", "Panel", "Fixed effects", "N", "R²"),
    align  = c("c","l","l","r","c","l","r","r"),
    escape = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("striped","hover","condensed","bordered"),
    full_width = TRUE, font_size = 13
  ) %>%
  column_spec(1, width = "2.5em", bold = TRUE, color = "#888") %>%
  column_spec(2, width = "9em",   bold = TRUE) %>%
  column_spec(3, width = "24em") %>%
  column_spec(4, width = "12em",  bold = TRUE, monospace = TRUE) %>%
  column_spec(5, width = "6em") %>%
  column_spec(6, width = "16em") %>%
  column_spec(7:8, width = "5em") %>%
  row_spec(c(3,4), background = "#FFF8EE") %>%
  row_spec(c(5,6), background = "#F0F7EE") %>%
  row_spec(2, extra_css = "border-bottom: 1px solid #bbb;") %>%
  row_spec(4, extra_css = "border-bottom: 1px solid #bbb;")

```

<p class="note">
**Notes.** Outcome: log(daily or monthly visits + 1), OLS throughout. All specifications cover Kano state only (Ungogo and Gabasawa LGAs, Aug 2024–Mar 2026). W1 has N = 24 (two sites × twelve months) — degrees of freedom are insufficient for reliable inference and the significant coefficient should not be read as a substantive finding; sign reversals appear in that panel on addition of further fixed effects. W6 is significant but reflects cross-LGA level differences between Gabasawa and Ungogo within each month-year cell rather than a within-unit temporal response; retained for transparency only. High R² values throughout are driven by fixed-effect absorption of seasonal and calendar variance, not by the weather predictors. &nbsp; &#42;&nbsp;p&nbsp;&lt;&nbsp;0.10 &nbsp; &#42;&#42;&nbsp;p&nbsp;&lt;&nbsp;0.05 &nbsp; &#42;&#42;&#42;&nbsp;p&nbsp;&lt;&nbsp;0.01
</p>

<hr>

### Table 3. &nbsp; Model details

```{r tab-details}

det <- tribble(
  ~Detail,           ~Z1,                     ~Z2,                   ~R3,                ~R4,                  ~W,
  "Specification",   "Primary",               "Strict",              "Full sample",      "Lag-time subset",    "Primary",
  "Outcome",         "ZD flag (penta, 12–23m)","No vaccine contact", "Confirmed return", "Confirmed return",   "log(visits + 1)",
  "Estimator",       "feglm logistic",        "feglm logistic",      "feglm logistic",   "feglm logistic",     "feols OLS",
  "Fixed effects",   "LGA",                   "LGA",                 "LGA",              "LGA",                "LGA, DOW, month-year",
  "Std. errors",     "Ward-clustered",        "Ward-clustered",      "LGA-clustered",    "LGA-clustered",      "Robust HC1",
  "<em>N</em>",      "69,366",                "69,366",              "40,236",           "16,132",             "1,216",
  "<em>R²</em>",     "0.061",                 "0.220",               "0.117",            "0.117",              "0.698"
)

det %>%
  kbl(
    col.names = c("", "(1)", "(2)", "(3)", "(4)", "(5)"),
    align     = c("l","c","c","c","c","c"),
    escape    = FALSE
  ) %>%
  kable_styling(
    bootstrap_options = c("hover","condensed","bordered"),
    full_width = TRUE, font_size = 13
  ) %>%
  add_header_above(
    c(" " = 1,
      "Zero-dose model" = 2,
      "Recovery model"  = 2,
      "Weather model"   = 1),
    background = c("white","#1D6FA4","#BA7517","#3A7D44"),
    color      = c("white","white","white","white"),
    bold       = TRUE
  ) %>%
  column_spec(1, bold = TRUE, width = "12em") %>%
  column_spec(2:6, width = "12em") %>%
  row_spec(5, extra_css = "border-bottom: 1px solid #ccc;") %>%
  row_spec(6:7, extra_css = "font-style: italic;")

```

<p class="note">
**Notes.** Columns (1) and (2) share the same predictor set; the strict outcome (2) requires no vaccine of any kind in facility records. Column (4) is restricted to tracing records with a prior facility visit date on file — this subset likely over-represents children with some prior system contact. The weather model R² of 0.698 is driven by day-of-week and month-year fixed effects; with LGA fixed effects only it is 0.088. Rimi LGA excluded from all specifications.
</p>

