---
output:
  pdf_document:
    latex_engine: xelatex
    toc: false
    number_sections: false
    fig_caption: true
    keep_tex: false
fontsize: 12pt
geometry: "left=2.5cm, right=2.5cm, top=2.5cm, bottom=2.5cm"
documentclass: article
header-includes:
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyhf{}
  - \fancyfoot[C]{\thepage}
  - \renewcommand{\headrulewidth}{0pt}
  - \usepackage{titling}
  - \pretitle{\begin{center}\LARGE\bfseries}
  - \posttitle{\end{center}}
  - \usepackage{setspace}
  - \onehalfspacing
  - \usepackage{etoolbox}
  - \AtBeginEnvironment{quote}{\itshape\small\singlespacing}
  - \AtEndEnvironment{quote}{\par}
---

```{r global-setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
source("_dissertation_setup.R")
```

\begin{titlepage}
\centering
\vspace*{2cm}

{\LARGE\bfseries What a Community-Based Digital Surveillance System Reveals --- and Conceals\par}
\vspace{0.6cm}
{\Large Zero-Dose Status, Tracing Effectiveness, and Weather in Northern Nigeria\par}

\vspace{2.5cm}
{\large University of Oxford\par}
{\large Nuffield Department of Medicine\par}
{\large MSc in International Health and Tropical Medicine\par}

\vspace{2.5cm}
{\large A dissertation submitted in partial fulfilment of the requirements\\ for the degree of MSc in International Health and Tropical Medicine\par}

\vspace{2.5cm}
{\large\bfseries Examination Candidate Number: [CANDIDATE NUMBER]\par}

\vspace{1cm}
{\large `r format(Sys.Date(), '%d %B %Y')`\par}

\vspace{1cm}
{\large\itshape Word count: 12,293\par}
{\small (inclusive of visualisation placeholders and figure/table blurbs; exclusive of abstract, table of contents, list of figures, references and appendices, per Section 2.4.2 of the IHTM Exam Conventions 2025--26)\par}

\vfill
\end{titlepage}

\newpage

\section*{Candidate's contribution to the project}

\textcolor{orange}{\textbf{[PLACEHOLDER --- replace before submission. Maximum 500 words per Section 2.4.2 of the Exam Conventions.]}}

Insert here exactly what you undertook on this project (analysis design, code, drafting) and what help you received and from whom, naming supervisors and collaborators as appropriate (e.g. academic adviser, placement supervisor, DATHARM staff who provided data access or field context), plus a brief statement of research ethics approval covering the MCHTrack data used.

\textcolor{orange}{\textbf{[ETHICS REFERENCE — insert approval reference/number here]}}

\newpage

\section*{Abstract}

\textbf{Background:} \textcolor{orange}{[placeholder --- 1--2 sentences framing the zero-dose problem and MCHTrack]}

\textbf{Methods:} This thesis uses four linked datasets from DATHARM's MCHTrack programme, which combines community-health-worker-led identification, referral and defaulter tracing across Kano and Katsina, Nigeria, answering three questions via ward-level fixed-effects logistic regression (zero-dose status, recovery through defaulter tracing) and a count-based weather model (daily facility visits against rainfall, heat and vegetation).

\textbf{Results:} Distance to a health facility is the most consistent predictor of zero-dose status, though its magnitude is highly sensitive to how zero-dose is defined (a non-significant $\sim$4\% higher odds per km under the primary definition vs. an 87\% higher odds per km under a strict, no-contact-at-all definition). Recovery through defaulter tracing is driven almost entirely by SMS/phone contact rather than distance, and varies sharply by how "recovered" is defined (42.6\% strict vs. 77.2\% permissive, pooled). No weather variable shows a significant association with either zero-dose status or facility visit volume, a null that survives offset, spline and negative-binomial robustness checks.

\textbf{Conclusion:} Routine administrative data of this kind is a genuinely useful but bounded evidence source: it reveals patterns survey-based literature cannot, but every finding is shaped by what the underlying system was built to record, and by what it silently leaves out.

\textcolor{orange}{\textit{(Structured abstract, maximum 300 words per Section 2.4.2 --- count and tighten before submission.)}}

\newpage

\tableofcontents
\listoffigures

\newpage

```{r child-body, child="_dissertation_body.Rmd"}
```
