---
title: "Lost and Found: Tracking Unvaccinated Children and the Search for a Weather Signal in Northern Nigeria"
subtitle: "MSc in International Health and Tropical Medicine, Nuffield Department of Medicine, University of Oxford"
author: "Examination Candidate Number: 1102514"
date: "`r {mo<-c('January','February','March','April','May','June','July','August','September','October','November','December'); y<-as.integer(format(Sys.Date(),'%Y')); if (y>2100) y<-y-543; sprintf('%d %s %d', as.integer(format(Sys.Date(),'%d')), mo[as.integer(format(Sys.Date(),'%m'))], y)}`"
output:
  word_document:
    toc: false
    number_sections: false
    fig_caption: true
    reference_docx: 11_word_reference.docx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
source("11_dissertation_setup.R")
```

::: {custom-style="TitleInfo"}
Word count: approximately 9,200 (Chapters I-V main text only; excludes front matter, abstract, references, tables, figures and appendices per exam conventions)
:::

```{=openxml}
<w:p><w:r><w:br w:type="page"/></w:r></w:p>
```

## Candidate's Contribution and Declaration

::: {custom-style="IntroBlock"}

This dissertation analyses programme data from MCHTrack, a community health worker-based digital immunisation tracking system operated by DATHARM, a Nigerian non-governmental organisation, across Kano and Katsina states in northern Nigeria between August 2024 and March 2026.

I undertook a six-week in-person placement with DATHARM at their offices in Abuja, Nigeria, during which I obtained the primary data and conducted fieldwork. I designed the analytical approach myself: I decided how to define zero-dose status and recovery, which variables to include, which model specification to treat as primary, and how to interpret every result. I built and maintained the analytical pipeline across eight R scripts, provided as supplementary material alongside a link to the project's GitHub repository and described script by script in Appendix B, with substantial AI assistance in writing, debugging, and refactoring the code itself, described in full in the AI use statement below.

I undertook two site visits to primary healthcare centres in the wider Abuja area for contextual grounding, and held informal conversations with DATHARM staff, including programme coordinators for Kano and Katsina, which informed the qualitative context presented in the discussion. These conversations were informal and observational; I did not apply a structured interview protocol or conduct independent qualitative analysis.

I received the following assistance during this project. Dr Prabin Dahal and Dr Aisha Adamu (primary supervisors), Dr Lucy Harrison, and a geospatial methods advisor provided academic supervision; my supervisors reviewed a complete first draft of the dissertation and gave written feedback. DATHARM's data manager provided the raw programme data and answered clarifying questions about specific data anomalies I identified during the audit. DATHARM's Chief Executive Officer, Amina Ahmad, is named as a collaborator on this work and provided programme context and review. A research partner, Ahmed Ismail, conducted a parallel and independently written analysis of delivery mode effectiveness using the same MCHTrack dataset; his analysis does not form part of this dissertation.

I obtained statistical advice from my supervisors during normal supervision meetings. I did not engage specialist statistical consultancy.

:::

### Research ethics

::: {custom-style="IntroBlock"}

This dissertation did not require a separate ethics review. DATHARM already held ethical approval for the MCHTrack programme itself, obtained through the relevant Nigerian regulatory channels as part of its routine public health programming in Kano and Katsina, and my access to its programme data was governed by data use agreements signed with DATHARM. I did not need a separate ethics review of my own for three reasons: local approval for the underlying project was already in place before I began; I had no contact with the beneficiary population at any stage of the research; and I had no access to identifiable data, since DATHARM provided the data to me pseudonymised. My research plan was reviewed as part of the placement proposal process through the MSc IHTM programme before the placement began, and no concerns requiring formal ethics referral were identified at that stage.

:::

### Statement on the use of artificial intelligence

::: {custom-style="IntroBlock"}

I used Claude (Anthropic) throughout both the analysis and writing phases of this project.

During the analysis phase, I used it to help write, debug, and refactor the R scripts behind this dissertation's pipeline: fixing bugs, restructuring code when supervisors' feedback required a different model specification, and building the figures and tables.

During the writing phase, I used it for drafting and restructuring prose in response to supervisor comments, for style and consistency editing, and for a systematic citation audit that checked every external source cited in this dissertation against the original text.

:::

```{=openxml}
<w:p><w:r><w:br w:type="page"/></w:r></w:p>
```

## Abstract

::: {custom-style="Abstract"}

**Background.** Community health workers in many low-income countries use digital tools to find children who have never been vaccinated, known as "zero-dose" children, and to bring back those who started but did not finish. One such tool is MCHTrack, run by DATHARM, a Nigerian non-governmental organisation, in northern Nigeria. This dissertation asks how well a programme's records can support the claims made about it.

**Methods.** I analysed MCHTrack's records for roughly 71,000 children in Kano and Katsina from August 2024 to March 2026, comparing outcomes under different definitions of zero-dose status and recovery after a missed dose, and testing whether rainfall, heat, or plant cover explained variation in facility visits.

**Results.** How zero-dose is defined changes what the data can reveal. Under MCHTrack's everyday operational definition, distance to the nearest facility barely predicts zero-dose status: each kilometre raises the odds a non-significant four per cent. Under a stricter definition, no vaccine contact at all, distance raises the odds 87 per cent, fifteen times as much. Recovery showed a similar pattern. Only 42.6 per cent of children followed up have a confirmed, verifiable vaccination record, rising to 77.2 per cent once self-reports are also counted, nearly doubling from one unrecorded verification step. Rainfall, heat and vegetation cover showed no comparable effect on facility visits.

**Conclusion.** A single figure from a programme's records can support two different conclusions about a child's risk or a programme's performance, depending on a definitional or verification choice most readers never see. Any reported figure should state which definition produced it and what remains unverified. Weather's null deserves the same scrutiny: it fits field reports that social barriers dominate, but the same recording gaps mean a hidden effect cannot be ruled out. This caution extends beyond MCHTrack, to any digital immunisation registry built the same way.

:::

```{=openxml}
<w:p><w:r><w:br w:type="page"/></w:r></w:p>
<w:sdt><w:sdtPr><w:docPartObj><w:docPartGallery w:val="Table of Contents" /><w:docPartUnique /></w:docPartObj></w:sdtPr><w:sdtContent><w:p><w:pPr><w:pStyle w:val="TOCHeading" /></w:pPr><w:r><w:t xml:space="preserve">Table of Contents</w:t></w:r></w:p><w:p><w:r><w:fldChar w:fldCharType="begin" w:dirty="true" /><w:instrText xml:space="preserve">TOC \o "1-3" \h \z \u</w:instrText><w:fldChar w:fldCharType="separate" /><w:t xml:space="preserve">Right-click and choose "Update Field" (or press Ctrl+A then F9) to generate the table of contents.</w:t><w:fldChar w:fldCharType="end" /></w:r></w:p></w:sdtContent></w:sdt>
<w:p><w:r><w:br w:type="page"/></w:r></w:p>
```

## List of Figures

| | |
|---|---|
| Figure 1.1 | MCHTrack footprint map across Kano and Katsina |
| Figure 2.1 | A child's path through MCHTrack |
| Figure 2.2 | Duplicate record set sizes, by state and table |
| Figure 2.2b | Monthly facility visit volume, by state |
| Figure 2.3 | Variance-to-mean ratio of daily facility visit counts, by LGA |
| Figure 2.4 | Baseline distributions of age, distance, and enrolment timing, by state |
| Figure 3.1 | Distance and age at registration, by number of vaccine doses recorded |
| Figure 3.2 | Sample construction waterfall |
| Figure 3.3 | Zero-dose model coefficients |
| Figure 3.5 | Tracing outcomes and timing |
| Figure 3.6 | Recovery model coefficients |
| Figure 3.7 | Strict versus permissive recovery rate, by state |
| Figure 3.8 | Daily visit diagnostics, Kano's Ungogo and Gabasawa LGAs |
| Figure 3.9 | Weather robustness coefficient plot |
| Figure D.1 | Age at tracing, by state |
| Figure D.2 | Correlation among each model's predictors and outcome |
| Figure D.3 | Mean monthly rainfall by LGA (CHIRPS) |
| Figure D.4 | Mean daytime heat by LGA (ERA5/UTCI) |
| Figure D.5 | Mean vegetation greenness (NDVI) by LGA |
| Figure E.1 | Ward-level residuals, ranked by magnitude |
| Figure E.2 | Ward-level residuals, mapped |
| Figure F.1 | Off-network share of recovered defaulters, by ward |

## List of Tables

| | |
|---|---|
| Table 2.1 | Baseline characteristics of enrolled children, by site |
| Table 2.2 | Key definitions used throughout this thesis |
| Table 3.1a | Zero-dose model — regression summary |
| Table 3.1b | Recovery model — regression summary |
| Table 3.2 | Weather model — main results |
| Table 3.3 | Weather model — offset and non-linearity robustness |
| Table 4.1 | This thesis against the two closest registry-based precedents |
| Table 4.2 | Convergence of quantitative findings and field observation |
| Table C.1 | Weather model — negative binomial and OLS compared |

```{=openxml}
<w:p><w:r><w:br w:type="page"/></w:r></w:p>
```

```{r child-body, child="11_dissertation_body.Rmd"}
```
